On multilevel constraint decomposition methods \\ for nonlinear and nonlocal variational inequalities

### 2021-07-04 paper/mcd2.tex
More text and better structure.

\begin{document}
\title[On MCD methods for nonlinear and nonlocal VIs]{On multilevel constraint decomposition methods \\ for nonlinear variational inequalities}

@@

We extend the original algorithms and theory presented in \cite{Tai2003} in three particular directions:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item We do not assume that the continuum VI problem arises from optimization of a scalar objective, and, as shown by examples in Section \ref{sec:vi}, this represents the removal of a symmetry assumption in \cite{Tai2003}.  Nonetheless we can prove convergence at essentially the same rate (Section \ref{sec:convergence}). % HOPE
\item In the multilevel case we make the observation that, if implemented appropriately, V(0,1) ``up-slash'' cycles are intrinsically more efficient than the V(1,0) ``down-slash'' cycles (Section \ref{sec:multilevel}).  This observation seems to be new for multilevel CD methods; compare the comments on V(1,0) and V(1,1) cycles in \cite{GraeserKornhuber2009,Tai2003}.  This preference for up-smoothing is, apparently, special to CD methods and does not arise in corresponding unconstrained subspace decomposition \cite{Xu1992} approaches.
\item In the multilevel and nonlinear case, we make the algorithms more practical by addressing efficient storage of intermediate quantities.  In particular we show results from a full approximation storage (FAS; \cite{Brandt1977}) version of the multilevel CD algorithm.
\end{enumerate}

@@

In one example we consider a case where the residual functional is nonlocal, that is, where it is not a partial differential operator at all.  Instead its values instead arise from the solution of Stokes problems for a layer of fluid on a substrate which forms the obstacle \cite{SayagWorster2013}, so the corresponding FE operator discretization is not sparse.  In this case we cannot prove coercivity but we can demonstrate near-optimal complexity in practice.  % HOPE

@@

When $f$ is monotone, VI \eqref{eq:vi} generalizes the problem of minimizing a convex function over $\cK$.  In fact, suppose $F:\cV \to \RR$ is lower semi-continuous and (G\^ateau) differentiable with continuous derivative $F':\cV \to \cV'$.  Then $F$ is convex if and only if $F'$ is monotone \cite[Proposition I.5.5]{EkelandTemam1976}, and if $F$ is convex then \eqref{eq:vi} holds for $f=F'$ and $g=0$ if and only if
\begin{equation}
u = \operatorname{arg-min}_{v\in\cK} F(v) \label{eq:minimization}
\end{equation}
\cite[Proposition II.2.1]{EkelandTemam1976}.  Note that Tai \cite{Tai2003} addresses optimization problem \eqref{eq:minimization}.

However, not all VIs correspond to optimization.  For example, consider the following steady advection-diffusion problem over $H_0^1(\Omega)$.  (We denote Sobolev spaces by $W^{k,p}(\Omega)$ if $k\ge 0$ and $p\ge 1$ \cite{Evans2010}, and $W^{k,2}=H^k$ in particular.)

\begin{example}  \label{ex:advectiondiffusion}  Suppose $\Omega \subset \RR^d$ is bounded and open with polygonal boundary.  Let $\cV = H_0^1(\Omega)$ and consider $\cK = \{v\ge 0\} \subset \cV$.  For $\bX \in L^\infty(\Omega)^d$ such that $\Div \bX=0$, and $u$ in $\cK$ let
\begin{equation}
\ip{f(u)}{v} = \int_\Omega \grad u \cdot \grad v\,dx - (\bX \cdot \grad u) v\,dx. \label{eq:nongradientexample}
\end{equation}
FIXME any $g\in\cV'$; POINT OUT THAT IS STEADY ADVECTION-DIFFUSION VI; MAKE ARGUMENT FOLLOWING \cite{Elmanetal2014}; see also \cite{Kirby2010}; IF $\bX=0$ THEN IT IS OPTIMIZATION BUT OTHERWISE NOT; $2$-coercive if outflow on Neumann; It is easy to see that $|\ip{f(u)}{v}| \le c \|u\|_{\cV} \|v\|_{\cV}$, thus that $f:\cK \to \cV'$ is continuous.
\end{example}

Other examples of VI problems which are not of optimization type are found in ice sheet models \cite{Calvoetal2002,JouvetBueler2012} and other geophysical fluids problems \cite{Bueler2021conservation}.

The VIs solved in this paper satisfy a stronger form of coercivity than \eqref{eq:coercive}.  This stronger assumption will be used in proving convergence of MCD methods.

@@

It is easy to see that if $f$ is $p$-coercive then it is monotone, strictly monotone, and coercive, and thus the following result, which states that the VI problems considered in this paper are well-posed, is immediate.

\begin{theorem}  \label{thm:viwellposed}  If $f:\cK \to \cV'$ is continuous and $p$-coercive (for $p>1$) then there exists a unique $u\in \cK$ so that VI \eqref{eq:vi} holds.
\end{theorem}

Our multilevel algorithms (next section) are based on both subspace and constraint decompositions, as follows.  Suppose there are $m<\infty$ subspaces $\cV_i \subset \cV$ so that
\begin{equation}
\cV = \sum_{i=0}^{m-1} \cV_i \label{eq:subspacedecomp}
\end{equation}
holds in the sense that if $w \in \cV$ then there exist $w_i \in \cV_i$ so that $w = \sum_i w_i$ \cite{Xu1992}.  Suppose further that $\cK_i \subset \cV$ are nonempty, closed, and convex subsets such that
\begin{equation}
\cK_i \subset \cV_i \qquad \text{and} \qquad \cK = \sum_{i=0}^{m-1} \cK_i. \label{eq:constraintdecomp}
\end{equation}
The constraint decomposition \cite{Tai2003} sum in \eqref{eq:constraintdecomp} is required to hold in two senses: \emph{(i)}~if $w \in \cK$ then there exist $w_i \in \cK_i$ so that $w = \sum_i w_i$, and \emph{(ii)}~if $z_i \in \cK_i$ for each $i$ then $\sum_i z_i \in \cK$.  (The latter sense is automatic in \eqref{eq:subspacedecomp} because $\cV_i$ are subspaces.)  Note that neither decomposition is unique (in general), and that $\cK_i \not\subset \cK$ in many applications.

Following \cite{Tai2003}, for each $\cK_i$ we also assume a (generally) nonlinear restriction operator $R_i : \cK \to \cK_i$ such that % SHOULD WE HAVE  R_i:\cV \to \cK_i?
\begin{equation}
v = \sum_{i=0}^{m-1} R_i v \qquad \text{if } v \in \cK.  \label{eq:constraintrestrictionsum}
\end{equation}
An \emph{constraint decomposition} of the constraint set $\cK$ is a choice of $\cV_i,\cK_i,R_i$ satisfying the above hypotheses.

In the next section we will introduce FE spaces and describe practical algorithms, but, as noted in the Introduction, constraint decompositions may be defined at the level of the continuum problem.  The following examples illustrate constraint decompositions of \emph{obstacle problems} \cite{GraeserKornhuber2009}; Example \ref{ex:domaindecomposition} is an overlapping domain decomposition and \ref{ex:frequencydecomposition} is a nonoverlapping frequency decomposition.  The latter example can be compared to the multilevel FE decomposition in the next section.

@@

\begin{pcode}[ht]
\begin{pseudo*}
\pr{cd-parallel}(f:\cK \to \cV', g\in \cV', \{\cK_i\}, \{R_i:\cK \to \cK_i\}, 0<\alpha\le 1, u\in\cK)\text{:} \\+
    for $i \in \{0,\dots,m-1\}$: \qquad \qquad \ct{parallel for} \\+
        $\hat w_i\in \cK_i$: \quad $\ip{f(u - R_i u + \hat w_i)}{v_i-\hat w_i} \ge g(v_i-\hat w_i) \quad \forall v_i\in \cK_i$ \\-
    return $w=(1-\alpha) u + \alpha \sum_i \hat w_i$
\end{pseudo*}
\caption{The basic parallel CD algorithm updates an iterate $u$ by solving VIs over each constraint set $\cK_i$ in any order, giving $w\in \cK$.}
\label{pc:cdparallel}
\end{pcode}

\begin{pcode}[ht]
\begin{pseudo*}
\pr{cd-sequential}(f:\cK \to \cV', g\in \cV', \{\cK_i\}, \{R_i:\cK \to \cK_i\}, 0<\alpha\le 1, u\in\cK)\text{:} \\+
    for $i = 0,\dots,m-1$: \qquad \qquad \ct{in order} \\+
        $\hat w_i\in \cK_i$: \quad $\displaystyle \ip{f\Big(\sum_{j<i} w_j + \hat w_i + \sum_{j>i} R_j u\Big)}{v_i-\hat w_i} \ge g(v_i-\hat w_i) \quad \forall v_i\in \cK_i$ \\
        $w_i = (1-\alpha) R_i u + \alpha \hat w_i$ \\-
    return $w= \sum_i w_i = (1-\alpha) u + \alpha \sum_i \hat w_i$
\end{pseudo*}
\caption{Each VI incorporates the previous solutions in this basic sequential CD algorithm; compare Pseudocode \ref{pc:cdparallel}.}
\label{pc:cdsequential}
\end{pcode}

In practice the space $\cV$ will be a finite element (FE) space because we will solve VI \eqref{eq:vi} over a mesh, and thus $\cV$ and each $\cV_i$ will be finite-dimensional.

\section{Multilevel constraint decomposition algorithms} \label{sec:multilevel}

@@

\section{Results} \label{sec:results}

@@

\small

\bigskip
\bibliography{mcd2}
\bibliographystyle{siam}

>>>>

\begin{document}
\title[On MCD methods for nonlinear and nonlocal VIs]{On multilevel constraint decomposition methods \\ for nonlinear and nonlocal variational inequalities}

@@

We extend the original algorithms and theory presented in \cite{Tai2003} in three particular directions:
\renewcommand{\labelenumi}{\emph{(\roman{enumi})}}
\begin{enumerate}
\item We do not assume that the continuum VI problem arises from optimization of a scalar objective, and, as shown by examples in Section \ref{sec:vi}, this represents the removal of a symmetry assumption in \cite{Tai2003}.  Nonetheless we can prove convergence at essentially the same rate (Section \ref{sec:convergence}). % HOPE
\item In the multilevel case we make the observation that, if implemented appropriately, V(0,1) ``up-slash'' cycles are intrinsically more efficient than the V(1,0) ``down-slash'' cycles (Section \ref{sec:multilevel}).  This observation seems to be new for multilevel CD methods; compare the comments on V(1,0) and V(1,1) cycles in \cite{GraeserKornhuber2009,Tai2003}.  This preference for up-smoothing is, apparently, special to CD methods and does not arise in corresponding unconstrained subspace decomposition \cite{Xu1992} approaches.
\item In the multilevel and nonlinear case, we make the algorithms more practical by addressing efficient storage of intermediate quantities.  In particular we show results from a full approximation storage (FAS; \cite{Brandt1977}) version of the multilevel CD algorithm (Section \ref{sec:resultslocal}).
\end{enumerate}

@@

In one example we consider a case where the residual functional is nonlocal, that is, where it is not a partial differential operator at all (Section \ref{sec:resultsnonlocal}).  Instead its values instead arise from the solution of Stokes problems for a layer of fluid on a substrate which forms the obstacle \cite{SayagWorster2013}, so the corresponding FE operator discretization is not sparse.  In this case we cannot prove coercivity but we can demonstrate near-optimal complexity in practice.  % HOPE

@@

The VIs solved in this paper satisfy a stronger form of coercivity than \eqref{eq:coercive}.\footnote{An exception is the problem in Section \ref{sec:resultsnonlocal}; there the functional $f$ is not known to be coercive.}

@@

It is easy to see that if $f$ is $p$-coercive then it is monotone, strictly monotone, and coercive, and thus the following result holds.  It states that the $p$-coercive VI problems considered in this paper are well-posed.

\begin{theorem}  \label{thm:viwellposed}  If $f:\cK \to \cV'$ is continuous and $p$-coercive (for $p>1$) then there exists a unique $u\in \cK$ solving VI \eqref{eq:vi}.
\end{theorem}

When $f$ is monotone, VI \eqref{eq:vi} generalizes the problem of minimizing a convex function over $\cK$.  In fact, suppose $F:\cV \to \RR$ is lower semi-continuous and (G\^ateau) differentiable with continuous derivative $F':\cV \to \cV'$.  Then $F$ is convex if and only if $F'$ is monotone \cite[Proposition I.5.5]{EkelandTemam1976}, and if $F$ is convex then \eqref{eq:vi} holds for $f=F'$ and $g=0$ if and only if
\begin{equation}
u = \operatorname{arg-min}_{v\in\cK} F(v) \label{eq:minimization}
\end{equation}
\cite[Proposition II.2.1]{EkelandTemam1976}.

Note that some numerical solvers for VIs, including the constraint decomposition methods of Tai \cite{Tai2003} (below), address optimization problem \eqref{eq:minimization} subject to the assumption that $F'$ is 2-coercive.  However, not all VIs correspond to optimization, and furthermore some optimization problems involve $p$-coercive residuals.  (The latter possibility is mentioned by \cite{Tai2003}; see also \cite{TaiXu2002}.)  We provide three examples.  Note that Sobolev spaces are denoted $W^{k,p}(\Omega)$ \cite{Evans2010}, with $W^{k,2}=H^k$ in particular.

\begin{example}  \label{ex:advectiondiffusion}  Consider a steady advection-diffusion problem over a bounded domain $\Omega \subset \RR^d$, with polygonal boundary, and suppose the boundary is partitioned $\partial\Omega = \partial_D\Omega \cup \partial_N\Omega$ with $\partial_D\Omega$ of positive measure.  Let $\cV = H_0^1(\Omega)$ be the space of functions with value zero on $\partial_D\Omega$; for simplicity suppose zero Dirichlet and Neumann data.  Consider a velocity field $\bX \in L^\infty(\Omega)^d$ such that $\Div \bX=0$, and define the functional
\begin{equation}
\ip{f(u)}{v} = \int_\Omega \eps \grad u \cdot \grad v - (\bX \cdot \grad u) v\,dx \label{eq:advectiondiffusion}
\end{equation}
for $\eps>0$.  Finally, suppose a nonnegativity constraint $\cK = \{v\ge 0\} \subset \cV$ on the solution \cite[for example]{ChangNakshatrala2017}, and consider VI \eqref{eq:vi} for any $g\in\cV'$.  It is easy to see that $|\ip{f(u)}{v}| \le (\eps + \|\bX\|_\infty) \|u\|_{\cV} \|v\|_{\cV}$, thus that $f:\cK \to \cV'$ is continuous.  FIXME  MAKE ARGUMENT FOLLOWING \cite{Elmanetal2014}; see also \cite{Kirby2010}; IF $\bX=0$ THEN IT IS OPTIMIZATION BUT OTHERWISE NOT; $2$-coercive if outflow on Neumann
\end{example}

\begin{example}  \label{ex:porousgeneralization}  FIXME
\begin{equation}
\ip{f(u)}{v} = \int_\Omega \phi(u) \grad u \cdot \grad v\,dx \label{eq:porousgeneralization}
\end{equation}
\end{example}

\begin{example}  \label{ex:plaplacian}  FIXME
\begin{equation}
\ip{f(u)}{v} = \int_\Omega |\grad u|^{p-2} \grad u \cdot \grad v\,dx \label{eq:plaplacian}
\end{equation}
\end{example}

Examples \ref{ex:advectiondiffusion} and \ref{ex:porousgeneralization} are VI problems which do not arise from optimization.  Further examples of such VIs are in ice sheet models \cite{Calvoetal2002,JouvetBueler2012} and other geophysical fluids \cite{Bueler2021conservation}.  Example \ref{ex:plaplacian} is an optimization-type problem of $p$-coercive type.  Numerical solver performance for Examples \ref{ex:advectiondiffusion}--\ref{ex:plaplacian} will be considered in section \ref{sec:resultslocal}.  % HOPE

Our multilevel algorithms (next section) are based on both subspace and constraint decompositions, as follows.  Suppose there are $m<\infty$ subspaces $\cV_i \subset \cV$ so that
\begin{equation}
\cV = \sum_{i=0}^{m-1} \cV_i \label{eq:subspacedecomp}
\end{equation}
holds in the sense that if $w \in \cV$ then there exist $w_i \in \cV_i$ so that $w = \sum_i w_i$ \cite{Xu1992}.  Suppose further that $\cK_i \subset \cV$ are nonempty, closed, and convex subsets such that
\begin{equation}
\cK_i \subset \cV_i \qquad \text{and} \qquad \cK = \sum_{i=0}^{m-1} \cK_i. \label{eq:constraintdecomp}
\end{equation}
The sum in \eqref{eq:constraintdecomp} is required to hold in two senses: \emph{(i)}~if $w \in \cK$ then there exist $w_i \in \cK_i$ so that $w = \sum_i w_i$, and \emph{(ii)}~if $z_i \in \cK_i$ for each $i$ then $\sum_i z_i \in \cK$.  (The latter sense is automatic in \eqref{eq:subspacedecomp} because $\cV_i$ are subspaces.)  Note that neither decomposition is unique (in general), and that $\cK_i \not\subset \cK$ in many applications.

Following \cite{Tai2003}, for each $\cK_i$ we also assume a (generally) nonlinear restriction operator $R_i : \cK \to \cK_i$ such that % SHOULD WE HAVE  R_i:\cV \to \cK_i?
\begin{equation}
v = \sum_{i=0}^{m-1} R_i v \qquad \text{if } v \in \cK.  \label{eq:constraintrestrictionsum}
\end{equation}
A \emph{constraint decomposition} (CD) of the set $\cK$ is a choice of $\cV_i,\cK_i,R_i$ satisfying the above hypotheses.

In the Section \ref{sec:multilevel} we will introduce discretizations and describe practical algorithms, but CDs may be defined at the level of the continuum problem.  The following examples illustrate CDs of obstacle problems \cite{GraeserKornhuber2009}.   Example \ref{ex:domaindecomposition} is an overlapping domain decomposition and Example \ref{ex:frequencydecomposition} is a nonoverlapping frequency decomposition.  The latter can be compared to the multilevel FE decomposition in Section \ref{sec:multilevel}.

@@

We now state the basic CD algorithms as iterations which solve VI problems over each set $\cK_i$.  (By contrast, the algorithms in \cite{Tai2003} involve iterative optimization over the $\cK_i$.)  There is a parallel (additive) form and a successive (multiplicative) form; these generalize the Jacobi and Gauss-Seidel iterations \cite{Greenbaum1997}, respectively.  Each algorithm is stated as a function which computes a new iterate $w\in\cK$ from a previous iterate $y \in \cK$, the result being an improved approximation of the solution $u\in \cK$ of VI problem \eqref{eq:vi}.

\begin{pseudo*}
\pr{cd-parallel}(\{\cK_i\}, \{R_i:\cK \to \cK_i\}, 0<\alpha\le 1, y\in\cK)\text{:} \\+
    for $i \in \{0,\dots,m-1\}$: \qquad \qquad \ct{parallel for} \\+
        $\hat w_i\in \cK_i$: \quad $\ip{f(y - R_i y + \hat w_i)}{v_i-\hat w_i} \ge g(v_i-\hat w_i) \quad \forall v_i\in \cK_i$ \\-
    return $w=(1-\alpha) y + \alpha \sum_i \hat w_i$
\end{pseudo*}

\begin{pseudo*}
\pr{cd-successive}(\{\cK_i\}, \{R_i:\cK \to \cK_i\}, 0<\alpha\le 1, y\in\cK)\text{:} \\+
    for $i = 0,\dots,m-1$: \qquad \qquad \ct{in order} \\+
        $\hat w_i\in \cK_i$: \quad $\displaystyle \ip{f\Big(\sum_{j<i} w_j + \hat w_i + \sum_{j>i} R_j y\Big)}{v_i-\hat w_i} \ge g(v_i-\hat w_i) \quad \forall v_i\in \cK_i$ \\
        $w_i = (1-\alpha) R_i y + \alpha \hat w_i$ \\-
    return $w = (1-\alpha) y + \alpha \sum_i \hat w_i$
\end{pseudo*}

Note that we assume a damping parameter $0<\alpha\le 1$.  The convergence results in the next section require substantial damping in the parallel algorithm, i.e.~$\alpha \le 1/m$, but the successive version can be shown to converge without damping ($\alpha=1$).  In Section \ref{sec:resultslocal} we demonstrate practical convergence for a larger range of $\alpha$ than suggested by the theory.  % HOPE

Observe that the expression $v_i - \hat w_i$ for $v_i,\hat w_i \in \cK_i$ appears in each algorithm, but, on the other hand, the reader may confirm that the argument to the residual functional $f$ is always an element of $\cK$.  To clarify this situation, note that by \eqref{eq:constraintrestrictionsum} one may write $v_i - \hat w_i$ as a difference of admissible vectors (i.e.~from $\cK$), namely
\begin{align*}
[y - R_i y + v_i] - [y - R_i y + \hat w_i] = \left[\sum_{j\ne i} R_j y + v_i\right] - \left[\sum_{j\ne i} R_j y + \hat w_i\right] &= v_i - \hat w_i, \label{eq:admissibledifference} \\
\left[\sum_{j<i} w_j + v_i + \sum_{j>i} R_j y\right] - \left[\sum_{j<i} w_j + \hat w_i + \sum_{j>i} R_j y\right] &= v_i - \hat w_i,  \notag
\end{align*}
for the two versions, respectively.  In other words, these algorithms require the approximate solution of VI \eqref{eq:vi} by looking among perturbations from the sets $\cK_i$.

FIXME key conceptual difference between the nonlinear and linear case; rewrite VI in \pr{cd-parallel} in the linear case; non-basic algorithm needed for $O(m)$ residual evaluation complexity in the nonlinear case (Section \ref{sec:multilevel})

@@

\section{Finite elements and multilevel constraint decomposition} \label{sec:multilevel}

Until now the algorithms can be interpreted as infinite-dimensional, e.g.~supposing constraint sets as defined in Examples \ref{ex:domaindecomposition} or \ref{ex:frequencydecomposition}.  However, in practice we will solve VI \eqref{eq:vi} over a mesh and the space $\cV$ will be a finite element (FE) space.  Thus $\cV$, and each subspace $\cV_i$, will be finite-dimensional.

FIXME triangulation

\section{Results for local variational inequalities} \label{sec:resultslocal}

@@

\section{Results for a nonlocal variational inequality} \label{sec:resultsnonlocal}

\small
\bibliography{mcd2}
\bibliographystyle{siam}
